Symmetric Circuits for Rank Gates

### 2017-12-07 sections/symmetric-circuits.tex
dec 7

A Boolean circuit $C$ computing a function from $\{0,1\}^n$ to $\{0,1\}$ is
usually described as a directed acyclic graph, with each gate $g$ that has $m$
incoming edges labelled with a Boolean function $f_g : \{0,1\}^m \rightarrow
\{0,1\}$ from a basis $\mathbb{B}$ and exactly $n$ input gates of $C$ are
labelled by variables. If the function $f_g$ can be arbitrary, then the circuit
$C$ must impose an order on the gates that provide the inputs to $g$. When we
say that $C$ is a directed acyclic graph, without further ordering the nodes, we
implicitly assume that the functions in the basis $\mathbb{B}$ are
\emph{symmetric}. That is to say $f_g$ is invariant under all permutations of
its $m$ inputs. This unstated assumption is pervasive in circuit complexity. In
particular, the standard Boolean basis of AND, OR and NOT gates as well as bases
with majority or threshold gates only contain symmetric functions.

@@

% AD - rewrote the paragraph to reduce the use of subscripts and superscripts.
Let $\tau = (R, S, \nu)$ be a many-sorted signature.
% with $R = \{R_1 \ldots, R_t\}$, $S = \{s_1, \ldots , s_p\}$, and $\nu(R_i) =
% (a^i_1, \ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for
% all $q \in [r_i]$.
Let $D := \biguplus_{s \in S} D_{s} = \{(s,d) : d \in D_s\}$, be a disjoint
union of non-empty sets, indexed by $S$, i.e.\ $D$ is potentially the universe
of a $\tau$-structure. We call $D$ a \emph{$\tau$-sorted set}. For each $P \in
R$, let $P^D$ denote the trivial relation over the many-sorted universe $D$,
i.e.\ $P^D := D_{a_1} \times \ldots \times D_{a_{r}}$, where $\nu(P) = (a_1,
\ldots , a_{r)})$. Let the \emph{index defined by the pair $(\tau, D)$} be the
set $\biguplus_{P\in R} P^{D}$. We use this set to index the input to a Boolean
function, as such we abuse notation and denote the index defined by $(\tau, D)$
by $\ind(\tau, D)$. We call $\tau$ the vocabulary and $D$ the universe of
$\ind(\tau, D)$.

For the remainder of this subsection we fix a vocabulary $\tau$ and a
many-sorted set $D=\biguplus_{s \in S} D_{s}$ and let $X := \ind (\tau, D)$. Let
$G \leq \sym_{X}$, and let $\sortsym(D) := \prod_{s \in S} \sym_{D_{s}}$. We
call the elements of $\sortsym(D)$ \emph{sorted permutations}. Note that
$\sortsym(D)$ can be identified with a subgroup of $\sym_X$. There is an obvious
action of $G$ on $X$. We can extend this to an action on the set of functions of
the form $f: X \rightarrow X$, for any set $H$, defined by $\sigma \cdot f(x) =
f(\sigma \cdot x)$ for all $x \in X$, $\sigma \in G$.

% If $H$ is surjective, we say that \emph{$f$ indexes $H$ by $X$} or \emph{$f$
% indexes $H$ by $\mathcal{D}$}.
We say that two functions $f, g : X \rightarrow H$ are \emph{$G$-equivalent} if
there exists $\sigma \in G$ such that $f = \sigma \cdot g$. We say $f$ and $g$
are \emph{isomorphism-equivalent} if they are $\sortsym(D)$-equivalent. We are
interested in the case where $H = \{0,1\}$, i.e.\ $f : X \rightarrow \{0,1\}$.
In this case $f$ defines for each relation $R^D_i$ a subset of that relation,
and thus determines a $\tau$-structure over the universe $D$. We note that two
functions $f, g: X \rightarrow \{0,1\}$ are isomorphism-equivalent if, and only
if, the two $\tau$-structures they determine are isomorphic.

@@

Structured functions give us a general framework for indexing the input sequence
to a Boolean function and associating with that function an appropriate symmetry
condition. Let $F$ be a $(\tau, D)$-structured function. We say $F$ is
\emph{$G$-invariant} if it is constant on $G$-equivalent input vectors, i.e.\
for all $G$-equivalent $\vec{x}, \vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) =
F(\vec{y})$. We say $F$ is \emph{isomorphism-invariant} if it is $\sortsym
(D)$-invariant. Note that a structured function with vocabulary $\tau$ is
isomorphism-invariant if, and only if, it decides a property of
$\tau$-structures.

>>>>

A Boolean circuit $C$ computing a function from $\{0,1\}^n$ to $\{0,1\}$ is
usually described as a directed acyclic graph, with each gate $g$ that has $m$
incoming edges labelled with a Boolean function $f_g : \{0,1\}^m \rightarrow
\{0,1\}$ from a basis $\mathbb{B}$ and exactly $n$ input gates of $C$ are
labelled by variables. If the function $f_g$ can be arbitrary, then the circuit
$C$ must impose an order on the gates that provide the inputs to $g$. When we
say that $C$ is a directed acyclic graph, without further ordering the nodes, we
implicitly assume that the functions in the basis $\mathbb{B}$ are
\emph{symmetric}. That is to say $f_g$ is invariant under all permutations of
its $m$ inputs. This unstated assumption is pervasive in circuit complexity. In
particular, the standard Boolean basis of $\AND$, $\OR$ and $\NOT$ gates as well as bases
with majority or threshold gates only contain symmetric functions.

@@

% AD - rewrote the paragraph to reduce the use of subscripts and superscripts.
Let $\tau = (R, S, \nu)$ be a many-sorted vocabulary.
% with $R = \{R_1 \ldots, R_t\}$, $S = \{s_1, \ldots , s_p\}$, and $\nu(R_i) =
% (a^i_1, \ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for
% all $q \in [r_i]$.
Let $D := \biguplus_{s \in S} D_{s} = \{(s,d) : d \in D_s\}$, be a disjoint
union of non-empty sets, indexed by $S$. We call $D$ a \emph{$\tau$-universe}.
For each $P \in R$, let $P^D$ denote the trivial relation over the many-sorted
universe $D$, i.e.\ $P^D := D_{a_1} \times \ldots \times D_{a_{r}}$, where
$\nu(P) = (a_1, \ldots , a_{r)})$. Let the \emph{index defined by the pair
  $(\tau, D)$} be the set $\biguplus_{P\in R} P^{D}$. We use this set to index
the input to a Boolean function, as such we abuse notation and denote the index
defined by $(\tau, D)$ by $\ind(\tau, D)$. We call $\tau$ the vocabulary and $D$
the universe of $\ind(\tau, D)$.

% GW: move sorted permutations into the background section
For the remainder of this subsection we fix a vocabulary $\tau$ and a
many-sorted set $D=\biguplus_{s \in S} D_{s}$ and let $X := \ind (\tau, D)$. Let
$G \leq \sym_{X}$, and let $\sortsym_D := \prod_{s \in S} \sym_{D_{s}}$. We
call the elements of $\sortsym_D$ \emph{sorted permutations}. Note that
$\sortsym(D)$ can be identified with a subgroup of $\sym_X$. There is an obvious
action of $G$ on $X$. We can extend this to an action on the set of functions of
the form $f: X \rightarrow X$, for any set $H$, defined by $\sigma \cdot f(x) =
f(\sigma \cdot x)$ for all $x \in X$, $\sigma \in G$.

% If $H$ is surjective, we say that \emph{$f$ indexes $H$ by $X$} or \emph{$f$
% indexes $H$ by $\mathcal{D}$}.
We say that two functions $f, g : X \rightarrow H$ are \emph{$G$-equivalent} if
there exists $\sigma \in G$ such that $f = \sigma \cdot g$. We say $f$ and $g$
are \emph{isomorphism-equivalent} if they are $\sortsym(D)$-equivalent. We are
often interested in the case where $H = \{0,1\}$, i.e.\ $f : X \rightarrow
\{0,1\}$. In this case $f$ defines for each relation $R^D_i$ a subset of that
relation, and thus determines a $\tau$-structure over the universe $D$. We note
that two functions $f, g: X \rightarrow \{0,1\}$ are isomorphism-equivalent if,
and only if, the two $\tau$-structures they determine are isomorphic.

@@

Structured functions give us a general framework for indexing the input sequence
to a Boolean function and associating with that function an appropriate symmetry
condition. Let $F$ be a $(\tau, D)$-structured function. We say $F$ is
\emph{$G$-invariant} if it is constant on $G$-equivalent input vectors, i.e. for
all $G$-equivalent $\vec{x}, \vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) =
F(\vec{y})$. We say $F$ is \emph{isomorphism-invariant} if it is $\sortsym
(D)$-invariant. Note that a structured function with vocabulary $\tau$ is
isomorphism-invariant if, and only if, it decides a property of
$\tau$-structures.
